Numerical simulations of the Cordilleran ice sheet through the last glacial cycle

### 2016-01-04 reviews/response-marshall.tex

\referee{%
    Several minor points and grammatical corrections are included in the
    attached text. Nothing that will require much thought -- this is a really
    impressive piece of research, overall, and I am hard-pressed to find any
    criticism of it. It is one of the easiest reviews I have ever done.
    Congratulations to the authors and thanks for this fine work.}

We have corrected grammatical mistakes given in the attachment. Thank you for
spotting them. And thank you again very much for supporting our manuscript with
such enthusiasm!

% ----------------------------------------------------------------------

\begin{thebibliography}{69}

\bibitem[{Cosma et~al.(2008)Cosma, Hendy, and Chang}]{Cosma.etal.2008}
Cosma, T., Hendy, I., and Chang, A.: Chronological constraints on Cordilleran
  Ice Sheet glaciomarine sedimentation from core MD02-2496 off Vancouver Island
  (western Canada), Quaternary Sci. Rev., 27, 941--955,
  \doi{10.1016/j.quascirev.2008.01.013}, 2008.

\bibitem[{Fisher et~al.(2008)Fisher, Osterberg, Dyke, Dahl-Jensen, Demuth,
  Zdanowicz, Bourgeois, Koerner, Mayewski, Wake, Kreutz, Steig, Zheng, Yalcin,
  Goto-Azuma, Luckman, and Rupper}]{Fisher.etal.2008}
Fisher, D., Osterberg, E., Dyke, A., Dahl-Jensen, D., Demuth, M., Zdanowicz,
  C., Bourgeois, J., Koerner, R.~M., Mayewski, P., Wake, C., Kreutz, K., Steig,
  E., Zheng, J., Yalcin, K., Goto-Azuma, K., Luckman, B., and Rupper, S.: The
  Mt Logan Holocene--late Wisconsinan isotope record: tropical Pacific--Yukon
  connections, The Holocene, 18, 667--677, \doi{10.1177/0959683608092236},
  2008.

\bibitem[{Fisher et~al.(2004)Fisher, Wake, Kreutz, Yalcin, Steig, Mayewski,
  Anderson, Zheng, Rupper, Zdanowicz, Demuth, Waszkiewicz, Dahl-Jensen,
  Goto-Azuma, Bourgeois, Koerner, Sekerka, Osterberg, Abbott, Finney, and
  Burns}]{Fisher.etal.2004}
Fisher, D.~A., Wake, C., Kreutz, K., Yalcin, K., Steig, E., Mayewski, P.,
  Anderson, L., Zheng, J., Rupper, S., Zdanowicz, C., Demuth, M., Waszkiewicz,
  M., Dahl-Jensen, D., Goto-Azuma, K., Bourgeois, J.~B., Koerner, R.~M.,
  Sekerka, J., Osterberg, E., Abbott, M.~B., Finney, B.~P., and Burns, S.~J.:
  Stable Isotope Records from Mount Logan, Eclipse Ice Cores and Nearby
  Jellybean Lake. Water Cycle of the North Pacific Over 2000 Years and Over
  Five Vertical Kilometres: Sudden Shifts and Tropical Connections, G\'{e}ogr.
  phys. Quatern., 58, 337, \doi{10.7202/013147ar}, 2004.

\bibitem[{Kienast and McKay(2001)}]{Kienast.McKay.2001}
Kienast, S.~S. and McKay, J.~L.: Sea surface temperatures in the subarctic
  northeast Pacific reflect millennial-scale climate oscillations during the
  last 16 kyrs, Geophys. Res. Lett., 28, 1563--1566,
  \doi{10.1029/2000GL012543}, 2001.

\bibitem[{Lisiecki and Raymo(2005)}]{Lisiecki.Raymo.2005}
Lisiecki, L.~E. and Raymo, M.~E.: A Pliocene-Pleistocene stack of 57 globally
  distributed benthic $\delta^{18}$O records, Paleoceanography, 20, PA1003,
  \doi{10.1029/2004pa001071}, 2005.

\bibitem[{M\"{u}ller et~al.(1998)M\"{u}ller, Kirst, Ruhland, von Storch, and
  Rosell-Mel{\'{e}}}]{Muller.etal.1998}
M\"{u}ller, P.~J., Kirst, G., Ruhland, G., von Storch, I., and
  Rosell-Mel{\'{e}}, A.: Calibration of the alkenone paleotemperature index
  $U^{K'}_{37}$ based on core-tops from the eastern South Atlantic and the
  global ocean (60$^\circ$N-60$^\circ$S), Geochim. Cosmochim. Ac., 62,
  1757--1772, \doi{10.1016/s0016-7037(98)00097-0}, 1998.

\bibitem[{Prahl and Wakeham(1987)}]{Prahl.Wakeham.1987}
Prahl, F.~G. and Wakeham, S.~G.: Calibration of unsaturation patterns in
  long-chain ketone compositions for palaeotemperature assessment, Nature, 330,
  367--369, \doi{10.1038/330367a0}, 1987.

\bibitem[{Prahl et~al.(1988)Prahl, Muehlhausen, and Zahnle}]{Prahl.etal.1988}
Prahl, F.~G., Muehlhausen, L.~A., and Zahnle, D.~L.: Further evaluation of
  long-chain alkenones as indicators of paleoceanographic conditions, Geochim.
  Cosmochim. Ac., 52, 2303--2310, \doi{10.1016/0016-7037(88)90132-9}, 1988.

\bibitem[{Seguinot et~al.(2014)Seguinot, Khroulev, Rogozhina, Stroeven, and
  Zhang}]{Seguinot.etal.2014}
Seguinot, J., Khroulev, C., Rogozhina, I., Stroeven, A.~P., and Zhang, Q.: The
  effect of climate forcing on numerical simulations of the {C}ordilleran ice
  sheet at the {L}ast {G}lacial {M}aximum, The Cryosphere, 8, 1087--1103,
  \doi{10.5194/tc-8-1087-2014}, 2014.

\bibitem[{Shackleton(1967)}]{Shackleton.1967}
Shackleton, N.: Oxygen Isotope Analyses and Pleistocene Temperatures
  Re-assessed, Nature, 215, 15--17, \doi{10.1038/215015a0}, 1967.

\bibitem[{Taylor et~al.(2014)Taylor, Hendy, and Pak}]{Taylor.etal.2014}
Taylor, M., Hendy, I., and Pak, D.: Deglacial ocean warming and marine margin
  retreat of the Cordilleran Ice Sheet in the North Pacific Ocean, Earth
  Planet. Sc. Lett., 403, 89--98, \doi{10.1016/j.epsl.2014.06.026}, 2014.

\bibitem[{Taylor et~al.(2015)Taylor, Hendy, and Pak}]{Taylor.etal.2015}
Taylor, M.~A., Hendy, I.~L., and Pak, D.~K.: The California Current System as a
  transmitter of millennial scale climate change on the northeastern Pacific
  margin from 10 to 50 ka, Paleoceanography, 30, 1168--1182,
  \doi{10.1002/2014pa002738}, 2015.

\end{thebibliography}
% ----------------------------------------------------------------------
% Interactive comment text ends
% ----------------------------------------------------------------------

### 2016-01-04 reviews/response-stumpf.tex

Thank you. We have added a reference to \citet{Stumpf.etal.2000} in the
discussion of deglacial flow directions (Sect.~4.3.3).

% ----------------------------------------------------------------------

\begin{thebibliography}{69}

\bibitem[{Ferbey et~al.(2013)Ferbey, Arnold, and Hickin}]{Ferbey.etal.2013}
Ferbey, T., Arnold, H., and Hickin, A.: Ice-flow indicator compilation, British
  Columbia., British Columbia Geol. Surv., Victoria, BC, Open-File 2013-06,
  \urlprefix\url{http://www.em.gov.bc.ca/Mining/Geoscience/PublicationsCatalogue/OpenFiles/2013/Pages/2013-06.aspx},
  2013.

@@

\bibitem[{Plouffe and Jett{\'{e}}(1997)}]{Plouffe.Jette.1997}
Plouffe, A. and Jett{\'{e}}, H.: Middle Wisconsinan sediments and paleoecology
  of central British Columbia: sites at Necoslie and Nautley rivers, Can. J.
  Earth Sci., 34, 200--208, \doi{10.1139/e17-016}, 1997.

\bibitem[{Stumpf et~al.(2000)Stumpf, Broster, and Levson}]{Stumpf.etal.2000}
Stumpf, A.~J., Broster, B.~E., and Levson, V.~M.: Multiphase flow of the late
  Wisconsinan Cordilleran ice sheet in western Canada, Geol. Soc. Am. Bull.,
  112, 1850--1863, \doi{10.1130/0016-7606(2000)112<1850:mfotlw>2.0.co;2}, 2000.

\bibitem[{Stumpf et~al.(2014)Stumpf, Ferbey, Plouffe, Clague, Ward, Paulen, and
  Bush}]{Stumpf.etal.2014}
Stumpf, A.~J., Ferbey, T., Plouffe, A., Clague, J.~J., Ward, B.~C., Paulen,
  R.~C., and Bush, A.~B.: Discussion: ``Streamlined erosional residuals and
  drumlins in central British Columbia, Canada'' by J. Donald {McClenagan},
  (2013) Geomorphology 189, 41--54, Geomorphology, 209, 147--150,
  \doi{10.1016/j.geomorph.2013.10.019}, 2014.

### 2016-01-05 reviews/response-jarosch.tex

% response-jarosch.tex
% ----------------------------------------------------------------------
\input{response-header}

\begin{document}
\textbf{Authors' response to A.~H.~Jarosch}
\bigskip

% ----------------------------------------------------------------------
% Interactive comment text begins
% ----------------------------------------------------------------------

\newcommand{\sechead}[1]{\bigskip\noindent\textbf{#1}}
\newcommand{\referee}[1]{\bigskip\textcolor{journalname}{\textit{#1}}}
\newcommand{\msquote}[1]{\begin{quote}\textit{#1}\end{quote}}
\newcommand{\doi}[1]{doi:\allowbreak\href{http://dx.doi.org/#1}{#1}}

@@

The aim of this manuscript, as we see it, is to create a bridge between two
scientific communities, which have long remained largely disconnected, by
presenting our work in such a way that both communities will understand it. We
refer here to the community of numerical ice sheet modellers on the one side,
and that of glacial geologists on the other. Thus from our viewpoint as
authors, the fundamental challenge you refer to is to communicate to each
community using their own terminology and methods to describe the results.

Because our manuscript is aimed at both communities, it is important for us, as
a team of co-authors with different backgrounds, that a balance is kept between
the description of the physics embedded in the numerical model, documented in
detail in many publications by PISM developers and users elsewhere
\citep[e.g.,][]{Bueler.Brown.2009, Winkelmann.etal.2011, Martin.etal.2011,
Aschwanden.etal.2013, Seguinot.2014, Bueler.Pelt.2015}, and the level of
regional detail in the discussion of geological evidence, documented elsewhere
as well though in a more fragmented way. We would be grateful if this point
could be taken into account when reading on.

Yet, while trying to keep this balance, we may have indeed omitted crucial
details related to the model set-up. We have made every attempt to correct this
in the new version of the manuscript.

\referee{%
    An overall sensitivity study of the parameters used in the model is
    completely lacking, thereby making it almost impossible to understand
    different responses of the ice sheet model. After reading the manuscript,
    one is left with the impression that the authors assume the PISM ice sheet
    model to be a black box which just requires one initial ``correct'' setup
    with literature values. This notion is reflected in the current manuscript,
    where almost all mismatches of model output with geological evidence (as
    discussed in section~4) are attributed to climate variations lacking in the
    proxy data, or climate-ice sheet feedback mechanisms not represented in the
    model chain. Similarly in a previous study Seguinot et~al. (2014) have
    focused only on the driving climate sensitivities and have omitted
    influences of the ice sheet model as well as mass balance model
    parameters even though they note in that study that these sensitivities
    require attention as well.}

@@

% ----------------------------------------------------------------------

\sechead{2 \quad Specific Comments}

\referee{%
    I refer to text locations in the discussion paper by page number (P) and
    line number range (L) for the specific comments.}

\referee{%
    \textbf{P4151 L11-16:} In this sentence the authors refer back to their
    previous work (Seguinot et~al., 2014) and highlight that the NARR
    temperature and precipitation fields are the most suitable present day
    climate datasets to be used. Especially since the NARR precipitation fields
    include steep precipitation gradients which are required as identified by
    et Seguinot et~al. (2014). NARR is delivered on a 32\,km Lambert grid,
    and thus it is questionable how ``steep'' these gradients can be, given
    the rather smooth representation of the existing topography on a 32\,km
    grid. Seguinot et~al. (2014) have partly discussed that however. NARR
    precipitation and temperature fields have been evaluated in detail based on
    available station data for large parts of the study domain dealt with in
    this manuscript. This evaluation (Jarosch et~al., 2012) demonstrated
    that NARR has difficulties simulating orographic processes in the Coast
    Mountains which in turn results in unrealistic atmospheric conditions over
    the Rocky Mountains. Jarosch et~al. (2012) further concluded that
    physics based downscaling is required to adequately drive glacier models in
    that region. The authors should argue in more detail here why they think
    that NARR precipitation fields at 32\,km are adequate to drive their model
    and reflect their arguments with the findings of Jarosch et~al. (2012).
    A solid argument here is of special importance as the authors assume the
    present day precipitation fields to be valid throughout their model time
    period (120\,ky to present) without further corrections (cf. section~2.4
    equation~6).}

@@

We admit that the NARR climate forcing used in our simulations has its
limitations.

In our previous study \citep{Seguinot.etal.2014}, we have evaluated the
performance of NARR in forcing constant-climate simulations of the Cordilleran
ice sheet against that of an observation-based data set
\citep[WorldClim]{Hijmans.etal.2005}. Indeed, the use of NARR in these
simulations produced slightly different patterns of glaciation relative to
WorldClim, including more extensive ice cover on the Columbia and Rocky
mountains \citep[Figs.~6--7]{Seguinot.etal.2014}. Our simulations have shown
that these differences are mainly caused by disparities in the precipitation
fields of the two data sets \citep[Figs.13--14]{Seguinot.etal.2014}, which may,
indeed, be related to the fact that the NARR has difficulties with simulating
orographic processes in some areas of steep topography as discussed by
\citet{Jarosch.etal.2012}.

In the light of your comments, we have introduced a discussion of the NARR and
its limitations in Sect.~2.6 (Climate forcing) and in Sect.~5.1.2 (Ice
configuration during MIS~2). We acknowledge that the implementation of the
downscaling method presented by \citet{Jarosch.etal.2012} may address these
limitations. However, extending this downscaling method to the entire the model
domain used in our study could be challenging, because the northern part of the
model domain is characterized by stronger precipitation gradients and includes
much fewer weather stations than the computational domain of
\citet{Jarosch.etal.2012}.

\referee{%
    \textbf{P4152 L11:}
    Basal topography is ``derived'' from ETOPO1 data. What does this mean? Do
    the authors just re-sample the DEM data to their 10\,km and 5\,km model
    grids (P4152 L21-22) or is there more processing done? The ETOPO1 data
    contains the present day ice volumes within the study region.
    Clarke et~al (2013)
    have estimated the ice volume in parts of that region to be $2530
    \pm 220\,km^2$, with maximum ice thicknesses up to 200\,m. It can be argued
    that the volume is negligible in this study and the authors should do so if
    they think this is appropriate, but I wonder about the ice thicknesses.
    Assuming that the authors did not remove the present day ice cover, basal
    topography could be up to 200\,m higher that it actually is in reality.
    Given their used temperature lapse rate of $6\,K\,km^{-1}$ (P4157 L1),
    parts of the Cordilleran ice sheet growing in those regions with 200\,m too
    high topography would experience a 1.2\,K colder atmosphere than it
    actually should in reality. This favours unrealistic ice growth and thus
    the omission of present day ice cover removal should be clearly argued for
    in the manuscript.}

@@

We have added a description of the limitations of ETOPO1 data in the methods,
and a comment on overestimated ice thicknesses in the Wrangell and St.-Elias
mountains in Sect.~5.2.2 (Major ice-dispersal centres).

\referee{%
    \textbf{P4153 L2-3:}
    That the ``shallow shelf approximation'' (SSA) is used as a ``sliding law''
    for the ``shallow ice approximation'' (SIA) is a confusing statement in
    this context. Bueler has coined the term in his 2009 paper as cited in the
    manuscript. However the casual reader will be confused at this point,
    especially since the authors state the pseudoplastic sliding law the model
    actually uses in equation~1. I would recommend to leave out the statement
    on the SSA being the ``sliding law'' for the SIA.}

We agree and have reformulated this statement. Thank you for spotting this.

\referee{%
    \textbf{P4153 L5-6:}
    As stated here, ice rheology within the used ice sheet model is based on
    Aschwanden et~al. (2012). This enthalpy based formulation has proofed
    itself to be very suitable for estimating ice rheology in ice sheet models,
    but it also depends on several parameters to translate enthalpy within the
    ice to ice viscosity (Aschwanden see et~al., 2012, equations 62-65).
    The authors do not mention any of these parameters (e.g. any of the rate
    factors or nonlinear power $n$) within the manuscript or in Table~1. I
    have mentioned above in the general comments section that parameters used
    in ice rheology and basal sliding formulations are important model
    parameters which will influence the ice sheet model output and that a basic
    sensitivity study on those parameters is required to understand the model
    results. Here the authors could start with listing the parameters used in
    the ice rheology formulation, than continue with estimating uncertainties
    for those from literature and afterwards perform additional model
    simulations to identify the influence of the chosen parameter sets on the
    ice volume and ice margin position history the model creates. In the end
    the authors will be able to identify the relative importance of
    uncertainties in driving climate as well as model parameters, which will
    strengthen their discussion in section~4.}

@@

The ``depth of 3\,km'' is measured the ice-bedrock interface downward. This is
where we apply geothermal heat flux as the lower boundary condition to the
bedrock thermal model. By ``computed subglacially'' we mean below the
ice-bedrock interface, into the bedrock. We have clarified these two points in
the text.

\referee{%
    \textbf{P4153 L16 - P4154 L0:}
    Here the authors describe the basal sliding setup in their model. However
    they do not explain how they came up with the parameters used in
    equations~1-3 that are listed in Table~1 (part on ``basal sliding''). What
    motivates these parameter choices (references?) and how sensitive is the
    model and its results to these choices? Both question come instantly to
    mind and need to be addressed in detail. Here a basic sensitivity study on
    how basal sliding parameters in the model control the outcome discussed in
    section~4 is in order and I strongly recommend to include one in the
    manuscript. The authors can start by estimating the uncertainties in the
    chosen basal sliding parameters and run two extra simulation runs with
    their preferred climate forcing and the end member values of the
    uncertainties. This would create the most simple sensitivity study with
    respect to basal sliding, but would be extremely helpful for the argument
    made above in my general comments.}

Our changes described above partly address this comment. We have added a new
column to Table~1 containing references motivating default parameter choices.

\referee{%
    \textbf{P4156 L3-5:}
    In addition to what I have stated above on the NARR precipitation fields
    and their suitability, it is important to state at this location in the
    manuscript how the 32\,km NARR data is translated to the 10 and 5\,km
    computational grids of the current study. I disagree with the notion that a
    32\,km precipitation field can be called ``high-resolution'' in the
    context of 10 and 5\,km grid based ice sheet modelling. The input data is
    either 3 or 6 times coarser than the numerical grid, thus not at all
    high-resolution. Seguinot et~al. (2014) state in their section~3.3 that
    the NARR data fields have been bilinearly interpolated to 10\,km resolution
    in their work. Did the authors do the same here for their 10 and 5\,km
    working grids? This is crucial information to be included in the
    manuscript. It has been demonstrated by spectral power analysis
    (Jarosch et~al., 2012) that the NARR precipitation fields do not contain
    any significant spacial information below approximately 39\,km resolution
    and that bilinear interpolation does not add any information whatsoever on
    smaller scales, which should come to no surprise. Physics-based downscaling
    techniques however are able to add spatial information to precipitation
    fields down to about 1\,km grid sizes (Jarosch et~al., 2012). Taking
    these findings into the current context of the manuscript at hand, the NARR
    precipitation fields can hardly be called ``high resolution'' with their
    effective precipitation grid size of 39\,km. The authors should argue for
    their choice of not performing any downscaling whatsoever to their
    computational grids of 10 and 5\,km for precipitation and temperature and
    discuss their choice in the light of the findings from
    Jarosch et~al. (2012). Temperature however is better constrained in
    NARR (Jarosch et~al., 2012) and contains spectral information down to
    10\,km resolution, which justifies the usage of NARR temperature fields on
    the 10\,km computational grid of this study. The 5\,km grid still needs to
    be argued for.}
